How many speech sounds are there in the world's languages?

### 2023-06-22 paper.Rmd
edits

The present study seeks to determine whether such a limit exists. To meet this goal, we propose several methods for modeling the relationship between sample size and number of linguistic types. We apply these methods to ask three core questions. First, how many phonemes are there in the world today (based on a limited sample)? Second, how many phonemes have there been in the history of human language? And lastly, how many phonemes are possible in human language? In Section 2, we motivate our research questions. In section 3, we describe the data and in Section 4 we present our methods and results. In Section 5, we discuss our results and avenues for future research.

@@

The question of whether there are limits on the number of distinct speech sounds that can appear in spoken languages has been recognized for some time. Alexander J. Ellis (1814--1890) stated as early as 1848 that the number of different speech sounds, i.e., candidate phonemes, is in principle infinite. At that time, teachers and philologists aimed to develop a transcription standard that could discriminate between all manners of speech sounds with orthographic symbols for purposes that included (particularly English) spelling reform and orthography development for newly documented and unwritten languages. But as practitioners increasingly converted the articulations of more and more languages into a universal phonetic alphabet, the need for new symbols also seemingly increased without limit [@Hockett1995-phoneme, pg. 63].^[This early work later became the basis for the International Phonetic Association (and its phonetic alphabet; both abbreviated IPA). First published in 1888, today the IPA provides and maintains a speech transcriptional device that encodes aspects of articulation, and acoustics, for both narrow (phonetic) and broad (phonemic) transcription. It is comprised of a set of alphabetic symbols and diacritics with the aim of representing all possible sounds in the world's spoken languages (there are extensions for disordered speech and speech pathology). The IPA provides a summary of linguists' phonetic knowledge. This system uses a set of "base" symbols, e.g., letters or combination of letters, that aim to encode phonemes, and the use of diacritics on base glyphs to provide phonetic detail. Its design also affords infinite transcription possibilities.] Finer and finer distinctions in phonetic detail created more and more space for novel phonemes to emerge. Thus, if the phonetic distinctions to be drawn are infinite, then tautologically so are the number of possible phonemes.

If there was any doubt about the infinite nature of the speech signal, instrumental phonetics put the nail in the coffin. Jean-Pierre Rousselot (1846--1924) studied acoustic properties of speech with the kymograph, originally developed to measure physiological patterns such as blood pressure. The later advent of the phonograph by Thomas Edison (1847--1931) in 1877 provided a means to record and reproduce sound. Then during WWII the sound spectogram was developed in secret to analyze ciphony by making sounds visible. The work on visible speech was released publicly after the war [@Potter1945] and the spectrogram became the staple of acoustic phonetic analysis beginning with @Joos1948 (Joos was the first linguist to have extensive access to the acoustic spectrograph). For speech, it measured fluctuations in air pressure, which capture duration, pitch, and intensity. The recorded speech waveforms showed visually that speech is dynamic and continuous, even if listeners perceive it as discrete, e.g., phonemes, syllables, words.^[This relationship between continuous and discrete is at the core of the debate about the interface between phonetics and phonology [@Whalen2019]. The division goes back to at least the late 19th century when phoneticians and dialectologists including Henry Sweet (1845--1912; 1877) and Jost Winteler (1846--1929; 1876) began making a distinction in language descriptions between sounds that change the meaning of words and those that do not.]

@@

The Zipfian distribution characterizes many aspects of language. Two features make it interesting for our purposes. First, it describes systems for which most types are rare, hence difficult to observe. Second, under certain conditions ($\alpha \ge 1$), the Zipf function is non-asymptotic: more types can always be added at the rank for singletons, and the expected type count computed. The fact that Maddieson's observations are consistent with the Zipfian distribution suggests that phoneme types are indeed difficult to discover (particularly at the fringe) and that there could be some room left for additional phonemes.

@@

```{r upsid_heaps, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}Cumulative segments in UPSID-451 follow Zipf's law"}
knitr::include_graphics("README_files/figure-gfm/zipf_distribution_upsid-1.png")
```

```{r phoible_heaps, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}: Cumulative segments in PHOIBLE (OIPL) follow Zipf's law"}
knitr::include_graphics("README_files/figure-gfm/zipf_distribution_phoible-1.png")
```

>>>>

The present study seeks to determine whether such a limit exists. To meet this goal, we propose several methods for modeling the relationship between sample size and number of linguistic types. We apply these methods to ask three core questions. First, how many phonemes are there in the world today (based on a limited sample)? Second, how many phonemes have there been in the history of human language? And lastly, how many phonemes are possible in human language? In Section 2, we motivate our research questions. In Section 3, we describe the data and in Section 4 we present our methods and results. In Section 5, we discuss our results and avenues for future research.

@@

The question of whether there are limits on the number of distinct speech sounds that can appear in spoken languages has been recognized for some time. Alexander J. Ellis (1814--1890) stated as early as 1848 that the number of different speech sounds, i.e., candidate phonemes, is in principle infinite. At that time, teachers and philologists aimed to develop a transcription standard that could discriminate between all manners of speech sounds with orthographic symbols for purposes that included (particularly English) spelling reform and orthography development for newly documented and unwritten languages (e.g., for bible translation). But as practitioners increasingly converted the articulations of more and more languages into a universal phonetic alphabet, the need for new symbols also seemingly increased without limit [@Hockett1995-phoneme, pg. 63].^[This early work later became the basis for the International Phonetic Association (and its phonetic alphabet; both abbreviated IPA). First published in 1888, today the IPA provides and maintains a speech transcriptional device that encodes aspects of articulation, and acoustics, for both narrow (phonetic) and broad (phonemic) transcription. It is comprised of a set of alphabetic symbols and diacritics with the aim of representing all possible sounds in the world's spoken languages (there are extensions for disordered speech and speech pathology). The IPA provides a summary of linguists' phonetic knowledge. This system uses a set of "base" symbols, e.g., letters or combination of letters, that aim to encode phonemes, and the use of diacritics on base glyphs to provide phonetic detail. Its design also affords infinite transcription possibilities.] Finer and finer distinctions in phonetic detail created more and more space for novel phonemes to emerge. Thus, if the phonetic distinctions to be drawn are infinite, then tautologically so are the number of possible phonemes.

If there was any doubt about the infinite nature of the speech signal, instrumental phonetics put the nail in the coffin. Jean-Pierre Rousselot (1846--1924) studied acoustic properties of speech with the kymograph, originally developed to measure physiological patterns such as blood pressure. The later advent of the phonograph by Thomas Edison (1847--1931) in 1877 provided a means to record and reproduce sound. Then during WWII the sound spectogram was developed in secret to analyze ciphony by making sounds visible. The work on visible speech was released publicly after the war [@Potter1945] and the spectrogram became the staple of acoustic phonetic analysis beginning with @Joos1948 (Joos was the first linguist to have extensive access to the acoustic spectrograph). For speech, it measured fluctuations in air pressure, which capture duration, pitch, and intensity. Thus, the ability to record, measure, and visualize speech shows that it is dynamic and continuous, even if listeners perceive it as discrete, e.g., phonemes, syllables, words.^[This relationship between continuous and discrete is at the core of the debate about the interface between phonetics and phonology [@Whalen2019]. The division goes back to at least the late 19th century when phoneticians and dialectologists including Henry Sweet (1845--1912; 1877) and Jost Winteler (1846--1929; 1876) began making a distinction in language descriptions between sounds that change the meaning of words and those that do not.]

@@

The Zipfian distribution characterizes many aspects of language (and other natural phenomena). Two features make it interesting for our purposes. First, it describes systems for which most types are rare, hence difficult to observe. Second, under certain conditions ($\alpha \ge 1$), the Zipf function is non-asymptotic: more types can always be added at the rank for singletons, and the expected type count computed. The fact that Maddieson's observations are consistent with the Zipfian distribution suggests that phoneme types are indeed difficult to discover (particularly at the fringe) and that there could be some room left for additional phonemes.

@@

```{r upsid_heaps, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}Cumulative segments in UPSID-451 follow Zipf's law"}
knitr::include_graphics("README_files/figure-gfm/heaps_upsid-1.png")
```

```{r phoible_heaps, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}: Cumulative segments in PHOIBLE (OIPL) follow Zipf's law"}
knitr::include_graphics("README_files/figure-gfm/heaps_phoible_oipl-1.png")
```

### 2023-06-22 paper.tex
edits

\title{How many speech sounds are there in the world's languages?}
\author{Steven Moran \& Nicholas A. Lester}
\date{21 June, 2023}

@@

The question of whether there are limits on the number of distinct
speech sounds that can appear in spoken languages has been recognized
for some time. Alexander J. Ellis (1814--1890) stated as early as 1848
that the number of different speech sounds, i.e., candidate phonemes, is
in principle infinite. At that time, teachers and philologists aimed to
develop a transcription standard that could discriminate between all
manners of speech sounds with orthographic symbols for purposes that
included (particularly English) spelling reform and orthography
development for newly documented and unwritten languages. But as
practitioners increasingly converted the articulations of more and more
languages into a universal phonetic alphabet, the need for new symbols
also seemingly increased without limit
(\protect\hyperlink{ref-Hockett1995-phoneme}{Hockett 1995}, pg.
63).\footnote{This early work later became the basis for the
  International Phonetic Association (and its phonetic alphabet; both
  abbreviated IPA). First published in 1888, today the IPA provides and
  maintains a speech transcriptional device that encodes aspects of
  articulation, and acoustics, for both narrow (phonetic) and broad
  (phonemic) transcription. It is comprised of a set of alphabetic
  symbols and diacritics with the aim of representing all possible
  sounds in the world's spoken languages (there are extensions for
  disordered speech and speech pathology). The IPA provides a summary of
  linguists' phonetic knowledge. This system uses a set of ``base''
  symbols, e.g., letters or combination of letters, that aim to encode
  phonemes, and the use of diacritics on base glyphs to provide phonetic
  detail. Its design also affords infinite transcription possibilities.}
Finer and finer distinctions in phonetic detail created more and more
space for novel phonemes to emerge. Thus, if the phonetic distinctions
to be drawn are infinite, then tautologically so are the number of
possible phonemes.

If there was any doubt about the infinite nature of the speech signal,
instrumental phonetics put the nail in the coffin. Jean-Pierre Rousselot
(1846--1924) studied acoustic properties of speech with the kymograph,
originally developed to measure physiological patterns such as blood
pressure. The later advent of the phonograph by Thomas Edison
(1847--1931) in 1877 provided a means to record and reproduce sound.
Then during WWII the sound spectogram was developed in secret to analyze
ciphony by making sounds visible. The work on visible speech was
released publicly after the war
(\protect\hyperlink{ref-Potter1945}{Potter 1945}) and the spectrogram
became the staple of acoustic phonetic analysis beginning with Joos
(\protect\hyperlink{ref-Joos1948}{1948}) (Joos was the first linguist to
have extensive access to the acoustic spectrograph). For speech, it
measured fluctuations in air pressure, which capture duration, pitch,
and intensity. The recorded speech waveforms showed visually that speech
is dynamic and continuous, even if listeners perceive it as discrete,
e.g., phonemes, syllables, words.\footnote{This relationship between
  continuous and discrete is at the core of the debate about the
  interface between phonetics and phonology
  (\protect\hyperlink{ref-Whalen2019}{Whalen 2016}). The division goes
  back to at least the late 19th century when phoneticians and
  dialectologists including Henry Sweet (1845--1912; 1877) and Jost
  Winteler (1846--1929; 1876) began making a distinction in language
  descriptions between sounds that change the meaning of words and those
  that do not.}

@@

The Zipfian distribution characterizes many aspects of language. Two
features make it interesting for our purposes. First, it describes
systems for which most types are rare, hence difficult to observe.
Second, under certain conditions (\(\alpha \ge 1\)), the Zipf function
is non-asymptotic: more types can always be added at the rank for
singletons, and the expected type count computed. The fact that
Maddieson's observations are consistent with the Zipfian distribution
suggests that phoneme types are indeed difficult to discover
(particularly at the fringe) and that there could be some room left for
additional phonemes.

@@

@@

>>>>

\title{How many speech sounds are there in the world's languages?}
\author{Steven Moran \& Nicholas A. Lester}
\date{22 June, 2023}

@@

The question of whether there are limits on the number of distinct
speech sounds that can appear in spoken languages has been recognized
for some time. Alexander J. Ellis (1814--1890) stated as early as 1848
that the number of different speech sounds, i.e., candidate phonemes, is
in principle infinite. At that time, teachers and philologists aimed to
develop a transcription standard that could discriminate between all
manners of speech sounds with orthographic symbols for purposes that
included (particularly English) spelling reform and orthography
development for newly documented and unwritten languages (e.g., for
bible translation). But as practitioners increasingly converted the
articulations of more and more languages into a universal phonetic
alphabet, the need for new symbols also seemingly increased without
limit (\protect\hyperlink{ref-Hockett1995-phoneme}{Hockett 1995}, pg.
63).\footnote{This early work later became the basis for the
  International Phonetic Association (and its phonetic alphabet; both
  abbreviated IPA). First published in 1888, today the IPA provides and
  maintains a speech transcriptional device that encodes aspects of
  articulation, and acoustics, for both narrow (phonetic) and broad
  (phonemic) transcription. It is comprised of a set of alphabetic
  symbols and diacritics with the aim of representing all possible
  sounds in the world's spoken languages (there are extensions for
  disordered speech and speech pathology). The IPA provides a summary of
  linguists' phonetic knowledge. This system uses a set of ``base''
  symbols, e.g., letters or combination of letters, that aim to encode
  phonemes, and the use of diacritics on base glyphs to provide phonetic
  detail. Its design also affords infinite transcription possibilities.}
Finer and finer distinctions in phonetic detail created more and more
space for novel phonemes to emerge. Thus, if the phonetic distinctions
to be drawn are infinite, then tautologically so are the number of
possible phonemes.

If there was any doubt about the infinite nature of the speech signal,
instrumental phonetics put the nail in the coffin. Jean-Pierre Rousselot
(1846--1924) studied acoustic properties of speech with the kymograph,
originally developed to measure physiological patterns such as blood
pressure. The later advent of the phonograph by Thomas Edison
(1847--1931) in 1877 provided a means to record and reproduce sound.
Then during WWII the sound spectogram was developed in secret to analyze
ciphony by making sounds visible. The work on visible speech was
released publicly after the war
(\protect\hyperlink{ref-Potter1945}{Potter 1945}) and the spectrogram
became the staple of acoustic phonetic analysis beginning with Joos
(\protect\hyperlink{ref-Joos1948}{1948}) (Joos was the first linguist to
have extensive access to the acoustic spectrograph). For speech, it
measured fluctuations in air pressure, which capture duration, pitch,
and intensity. Thus, the ability to record, measure, and visualize
speech shows that it is dynamic and continuous, even if listeners
perceive it as discrete, e.g., phonemes, syllables, words.\footnote{This
  relationship between continuous and discrete is at the core of the
  debate about the interface between phonetics and phonology
  (\protect\hyperlink{ref-Whalen2019}{Whalen 2016}). The division goes
  back to at least the late 19th century when phoneticians and
  dialectologists including Henry Sweet (1845--1912; 1877) and Jost
  Winteler (1846--1929; 1876) began making a distinction in language
  descriptions between sounds that change the meaning of words and those
  that do not.}

@@

The Zipfian distribution characterizes many aspects of language (and
other natural phenomena). Two features make it interesting for our
purposes. First, it describes systems for which most types are rare,
hence difficult to observe. Second, under certain conditions
(\(\alpha \ge 1\)), the Zipf function is non-asymptotic: more types can
always be added at the rank for singletons, and the expected type count
computed. The fact that Maddieson's observations are consistent with the
Zipfian distribution suggests that phoneme types are indeed difficult to
discover (particularly at the fringe) and that there could be some room
left for additional phonemes.

@@

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/heaps_upsid-1}

@@

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/heaps_phoible_oipl-1}
